---
title: "Analyzing GSS Data"
author: "Matt Lawrence"
date: "April 29, 2020"
output: pdf_document
keep_tex: true
---

## Welcome to R Studio!

Some navigation before we begin:

- The upper left pane is the *text editor* where you keep all of your commands.
- The lower left pane is the *console* where the commands will be executed.
- The upper right pane's `Environment` window will show the data that you load.
- The lower right pane will show previews of your R Markdown documents, plots, help documents, packages, and other features.

This file is an R Notebook which is an R Markdown file that allows you to integrate text, R code, and R output into a single document. For our purposes - and for producing transparent and reproducible research - the advantage of an R Notebook is that you can easily include text, notes, and comments that are not code (like this paragraph).

How does R know what is code and what is regular text? In an R Notebook, code goes in a fenced code block. To open the fence, type three backticks and an r in curly brackets on a single line; to close the fence, type three backticks on a single line like this:

```{r}

```

Everything between those fences will be run as code (unless it is preceded by a hashtag) when you press the green arrow in the code block.

We start by loading the data and packages. Note that you may need to install the packages.

```{r, error = FALSE, warning = FALSE, message = FALSE}

library(tidyverse) 
# This loads the tidyverse package

datalink <- "https://raw.githubusercontent.com/mjclawrence/soci301/s20/data_workshop/gss_clean.Rda"

load(url(datalink))

gss <- filter(gss, TRUE)
```


For an overview of the types of variables we have in our dataset, look to the top right pane and click the blue button with the white arrow next to the data frame name. There are two types of variables in our dataset. The `Int` variables are *integer* variables which means they are numeric. The `Factor` variables are categorical variables; their *levels* are the categories.

## Variables

What are the variables in this GSS subset (pulled from the 1996-2018 survey years)? Here are the variable names and labels. For further details about the wording of the questions, visit http://gssdataexplorer.norc.org

`abany` = Abortion if woman wants for any reason

`age` = Age in years

`anyhispanic` = Self-identified ethnic origin includes Hispanic (created from `hispanic`)

`class` = Subjective class identification

`conmedic` = Confidence in medicine

`conschls` = Confidence in schools and education system

`degree` = Highest degree completed

`disabled` = Respondent has a disability (created from `disabld1` - `disabld5`)

`health` = Self-reported health condition

`memnum` = Number of memberships

`natdrugy` = Are we spending too much, too little, or about the right amount on drug rehabilitation?

`polinter` = Expressed political views on internet past year

`pubsch` = All children attend public elementary and high school

`race` = Big race categories (white, black, other)

`racehisp` = Created from `race` and `hispanic` variables. If `anyhispanic` = Yes, rachisp = Hispanic.

`sellbaby` = Surrogate mothers should be permitted

`sex` = Respondent's sex

`sports30` = In the past 30 days, how often have you visited a web site for sports?

`tvhours` = Hours per day watching tv

`usmixed` = What percentage of population is of mixed heritage?

`year` = Survey year


## Distributions of Single Variables

Let's look at the distribution of the `age` variable. We can do this with the `summary()` function. Note that in the chunk below, we state the function to use and then in parentheses state the data frame name followed by a dollar sign and the variable name. This is the standard syntax for most R commands.

```{r}
summary(gss$age)
```

Characteristics you might reference to describe variables include the range (the minimum and maximum), the median (the middle point in the distribution), and the mean (the average). How would you describe this distribution? 

Notice that this summary distribution includes 103 observations with NA values. That tells us that 103 observations are missing data for the age variable. For some functions, we will have to manually remove the NAs from our analyses for them to run properly. For example, if we want to find the mean age, we add the `na.rm = TRUE` option to remove the NAs.

```{r}
mean(gss$age, na.rm=TRUE)
```



## Creating Categories

The values of the age variable are single years of age (18, 19, 20, etc.) But it might make sense to create categories for this variable. There are several ways to do this in R, some more efficient than what we will do below. But this is the most straightforward way to be able to easily amend the code for your own purposes in the future. In this chunk, we are creating a new variable in the `gss` data frame called `age_category` which divides the exisitng values of the `age` variable into the categories we previously saw the GSS use for their key trends analyses. We put the conditions for each category in brackets (this is called *indexing*). To the right of the arrow, we give the name we want to use for the new category level.

```{r}
gss$age_category[gss$age<35] <- "18-34" # use < for "less than"
gss$age_category[gss$age>=35 & gss$age<50] <- "35-49" # use >= for "greater than or equal to"
gss$age_category[gss$age>=50 & gss$age<65] <- "50-64"
gss$age_category[gss$age>=65] <- "65+"
gss$age_category[gss$age=="NA"] <- "NA" # use "==" for equal to
```

A categorical variable does not have the same kind of distribution as a numerical variable. Specifically, it does not have a mean or a median in the same way, so we would not use summary() with a categorical variable. Instead, let's look at a table to see the number of observations in each age category:

```{r}
table(gss$age_category)
```

What if we wanted proportions rather than frequencies? Wrap the table() in prop.table():

```{r}
round(prop.table(table(gss$age_category)),3)
```

Try making a proportion table for the `health` variable, which includes responses to the question: "Would you say your own health, in general, is excellent, good, fair, or poor?"

### REPLACE THIS LINE WITH YOUR CODE

```{r}
prop.table(table(gss$health))
```

There are a lot of decimal places in this proportion table. We can use the `round()` function to simplify. Note the `,3` option tied to the parentheses closing the `round()` function. That 3 asserts we want the proportions rounded to 3 decimal places.

```{r}
round(prop.table(table(gss$health)),3)
```


### Cross Tabulations

It is often helpful to know how one variable's distribution varies according to another variable (or variables). For example, we may want to see the proportion of respondents in each `degree` category who are in each `health` category. To do that, use the same code as above but include both variables (separated by a comma). The first variable should be the *row* variable and the second variable should be the *column* variable. It is standard to use the rows for your independent variable (or X) and the columns for your dependent variable (or Y).

```{r}
round(prop.table(table(gss$degree, gss$health), 1),3)
```

We added the `, 1` after the table because we want our row proportions to add up to 1 (and the row variable is our first variable). For column proportions, change the 1 to a 2.

```{r}
round(prop.table(table(gss$degree, gss$health), 2), 3)
```

If you do not include a 1 or 2, you will get cell proportions - the proportion of all observations in each cell. That is rarely helpful.

```{r}
round(prop.table(table(gss$degree, gss$health)),3)
```

What proportion of respondents in each age category have expressed political views on the internet in the last year (the `polinter` variable)?

### REPLACE THIS LINE WITH YOUR CODE

```{r}
round(prop.table(table(gss$age_category, gss$polinter),1),3)
```

Let's see how the age distribution in internet political usage varies by sex. Once you are working with at least three variables, it is more efficient to use the `group_by` and `summarize` functions to calculate measures for specific combinations of categorical levels.

```{r}
political_sex_table <- gss %>%
     group_by(age_category, sex) %>%
     filter(age_category!="NA") %>% # Use filter to restrict observations
     summarize(political_internet = 
                       mean(polinter=="Have done it in the past year",
                            na.rm=TRUE))

political_sex_table
```

## From Tables To Figures


### Column Plots

The information in the table we just made might be better conveyed through a figure. We will use the `ggplot2` package for figures. The foundation for ggplot is to think of a figure as having multiple layers. The bottom layer is where we name the object, state the dataset we are using, and then give some aesthetic mapping information. In this example, we will use the data in the instagram_sex_table created above, will use the values in the age_category column for our x-axis, will use the values in the used_instagram column for our y-axis, and will want different colored columns for males and females (the two values of the sex variable).

```{r}
political_sex_plot <- ggplot(political_sex_table, aes(x = age_category,
                                        y = political_internet,
                                        fill = sex))

political_sex_plot
```

Once we set the background for our plot, we add another layer to it. We will have a separate column for each sex in every age category, and the height of the column will be the proportion in that sex/age combination who have reported expressing themselves politically on the internet in the last year. 

```{r}
political_sex_plot + geom_col(position = "dodge")
```

Changing "dodge" to "fill" in the chunk above will plot the proportion in each category who have expressed themselves online who are female and male. That is not as useful for this analysis but could be helpful for other analyses.

Clean up the dodged figure by adding new axis titles, adding a plot title and subtitle, removing the title for the legend (the key for the "fill" variables), moving the legend to the bottom of the figure, and adding text labels for each column showing the proportion expressing themselves online for that age/sex combination.

```{r}
political_sex_plot + geom_col(position = "dodge") +
     labs(x = "Age",
          y = "Proportion",
          title = "Expressed Political Opinion On Internet In Last Year, By Age",
          subtitle = "2014 General Social Survey",
          fill = "") +
     theme(legend.position = "bottom") +
     geom_text(aes(label = round(political_internet,3)),
               position = position_dodge(1),
               vjust = 0)
```

### Historical Trend Lines

The last figure we will make is a historical trend line showing racial differences in responses to the courts variable. We start by finding, for each year and race combination, the proportion with each response to the courts question. We'll use the `racehisp` variable (created from the `race` and `anyhispanic` variables) here.

```{r}
conmedic_data <- gss %>%
     group_by(year, racehisp) %>%
     filter(racehisp!="NA") %>%
     summarize(prop_hardly_any = 
                    mean(conmedic == "Hardly any", na.rm=TRUE),
               prop_only_some = 
                    mean(conmedic == "Only some", na.rm=TRUE),
               prop_great_deal = 
                    mean(conmedic == "A great deal", na.rm=TRUE))
```

Next we reshape the data so there is one row for each year/race/confidence combination and all the proportions are in one column. We will also fix the order of the courts categories (with the `levels` option) and give them real names (with the `labels` option).

```{r}
conmedic_data <- conmedic_data %>%
     gather(category, proportion, prop_hardly_any:prop_great_deal) %>%
     mutate(category = factor(category,
                              levels = c("prop_hardly_any", 
                                         "prop_only_some", 
                                         "prop_great_deal"),
                              labels = c("Hardly any",
                                         "Only some",
                                         "A great deal")))
```

Now we are ready to build the plot. We need points and lines (rather than columns) this time, and we want the points and lines to be in different colors for each courts category. The `facet_grid` option will create a separate plot for each category of the race variable. We will also clean up the labels and legend again.

```{r}
conmedic_plot <- ggplot(conmedic_data, aes(x = year, y = proportion, 
                                       color = category, group = category))

conmedic_plot + geom_point() + geom_line() + facet_grid(.~racehisp) +
     labs(x = "Year",
          y = "Proportion",
          title = "Racial Differences in Confidence in Medicine",
          subtitle = "General Social Survey, 2000-2018",
          color = "") +
     theme(legend.position = "bottom")
```
